Evolution of ORM Systems: Formal Foundations

### 2013-01-14 doc/description.tex
added pkValue

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete values of concrete columns. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, ps)) = ps \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
		\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	 l \in Label, ps = Pair*, i \lhd Row
\end{align*}
The value stored in concrete column of the row is represented by pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		\inference{r = \varnothing}{\varnothing} \\
 		\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \in Rows, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

@@

@@

\paragraph{Insert Value} The function inserts a new pair into an existing row.
\begin{align*}
	\mathbf{insertValues} : Row \times Pair* \times Database \rightarrow Daatabase \\
	insertValues(r, p, d) = \begin{cases}
 		\inference{|p| > 1}{insertValues(r, head(p), d)\bullet insertValues(r, tail(p), d)}  \\ \\
 		\inference{|p| = 1 \wedge \forall e \in pairs(r) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{pairs(r) \cup p} \\ \\
	 	 Inconsistent \; d
 		\end{cases} \\
r \lhd Row, p = Pair*,d \lhd Databases
\end{align*}
\paragraph{Copy Value} The function copy values from one column in a table to another column in the same or in another table - according to mapping function $f_m$. 
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc}}{\begin{gathered}
			insertValues(second(p), \\ [l_{tc},value(pairByRef(l_{sc}, first(p)))],d)
		\end{gathered}}
 	\end{cases} \\
 	l_s, l_t, l_{sc}, l_{tc} \in Label, d \lhd Database
\end{align*}

@@

%\begin{align*}
%	v(l, fm, d) = \begin{cases}
%	\inference{fm = \varnothing}{d}\\ \\
%	\inference{|fm| > 1}{v(l, head(fm)) \bullet v(l, tail(fm))}\\ \\
%	\inference{|fm| = 1 \exists c \in pairs(first(fm)) : reference(c) = l \\ \wedge \exists e \in pairs(second(fm)) : reference(e) = l}{e \cup pairByRef(l,c)} 
% \end{cases}
%\end{align*}

>>>>

\paragraph{Stored Data} A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete values of concrete columns. Each row contains a name of a table it belongs to.
\begin{align*}
	\mathbf{Row} = < refTable, pkValue, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, n, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, n, ps)) = ps \\ \\
	\mathbf{pkValue} : Row \rightarrow N \\
	pkValue(row(l, n, ps)) = n \\ \\
	 l \in Label, ps = Pair*, n \in N
\end{align*}
The value stored in concrete column of the row is represented by the pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
The definition of data allows definition of function $contains$ which verify if the table contains data in context of concrete database and $instantiated$, which search for instances of a class with given name and returns true if there are rows for the table in the database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
\begin{align*}
	\mathbf{tableData} : Label \times Row* \rightarrow Row* \\
	tableData(l, r) = \begin{cases}
 		\inference{r = \varnothing}{\varnothing} \\
 		\inference{|r| > 1}{tableData(head(r)) \cup tableData(tail(r))}\\ \\
 		\inference{|r| = 1 \wedge refTable(r) = l}{r}\\\\
 		\inference{|r| = 1 \wedge refTable(r) \neq l}{\varnothing}
 \end{cases} \\ \\
	r \in Rows, l \in Label
\end{align*}
\begin{align*}
	\mathbf{instantiated} : Class \times Software \rightarrow Boolean \\
	instantiated(c, s) = \begin{cases} \inference{\exists t \in tables(database(s)) : name(t) = name(c) \\ \wedge contains(t, database(s))}{true} \\ \\
  false
 \end{cases} \\ \\
	 c \in classes(application(s)), s \lhd Software 
\end{align*}

@@

\paragraph{Find pair}
\begin{align*}
\mathbf{findPair} : Label \times Row \rightarrow Pair \\
	findPair(l, r) = \begin{cases}
		\inference{\exists p \in pairs(r) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
		l \in Label, r \lhd Row
\end{align*}

@@

\paragraph{Insert Value} The function inserts a new pair into an existing row.
\begin{align*}
	\mathbf{insertValues} : Row \times Pair* \times Database \rightarrow Daatabase \\
	insertValues(r, p, d) = \begin{cases}
 		\inference{|p| > 1}{insertValues(r, head(p), d)\bullet insertValues(r, tail(p), d)}  \\ \\
 		\inference{|p| = 1 \wedge \forall e \in pairs(r) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{pairs(r) \cup p} \\ \\
	 	 Inconsistent \; d
 		\end{cases} \\
r \lhd Row, p = Pair*,d \lhd Databases
\end{align*}
\paragraph{Copy Value} The function copy values from one column in a table to another column in the same or in another table - according to mapping function $f_m$. 
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc}}{\begin{gathered}
			insertValues(second(p), \\ [l_{tc},value(findPair(l_{sc}, first(p)))],d)
		\end{gathered}}
 	\end{cases} \\
 	l_s, l_t, l_{sc}, l_{tc} \in Label, d \lhd Database
\end{align*}

@@

%\begin{align*}
%	v(l, fm, d) = \begin{cases}
%	\inference{fm = \varnothing}{d}\\ \\
%	\inference{|fm| > 1}{v(l, head(fm)) \bullet v(l, tail(fm))}\\ \\
%	\inference{|fm| = 1 \exists c \in pairs(first(fm)) : reference(c) = l \\ \wedge \exists e \in pairs(second(fm)) : reference(e) = l}{e \cup findPair(l,c)} 
% \end{cases}
%\end{align*}
